---
title: "Rmarkdown exercise - putting it all together"
author: "rob"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For this exercise, we're going to put everything (exploratory data
analysis, visualization, and wrangling) together.

Part 1 will include some specific tasks you will be asked to complete as
a group.

Part 2 will ask you to select an aspect of the data that you find
interesting (a question you have about the data, perhaps) and report on
this question using summary statistics and data visualization.

## Data

The data we're going to be exploring is treatment data from 20
individuals who received Sound Production Treatment for Post-stroke
Apraxia of Speech from Julie Wambaugh and colleagues (2017).

Here's a quick summary of the experimenal design, exerpted from a recent
preprint:

> Wambaugh et al., (2017) reported the effects of Sound Production
> Treatment for apraxia of speech and aphasia for 20 individuals in a
> multiple-baseline design under two experimental conditions: blocked
> and random practice. Sound Production Treatment uses "therapeutic
> techniques of modeling and repetition, contrastive practice,
> orthographic cueing, integral stimulation, and articulatory cueing" in
> a "response-contingent hierarchy" to target phoneme production (p.
> 1744). The expectation is that repeated practice based on principles
> of motor learning will improve the production of target phonemes in
> the treated context, and (ideally) generalize to the production of
> those phonemes in words that are not explicitly treated.

> In Wambaugh et al., (2017), participants received treatment on two
> lists, one in each experimental condition (random and blocked). Each
> list contained treated items and untreated generalization items for
> two target phonemes. Items consisted of a single word or occasionally
> a 2-3 word phrase. For 16/20 participants, treated lists contained 20
> items (10 for each phoneme), and untreated lists contained 10 items (5
> for each phoneme). For these participants, accuracy was determined
> based on the production of the target phoneme within the item. For
> 4/20 participants, treated lists contained 10 items (5 for each
> phoneme), and untreated lists contained 6 items (3 for each phoneme).
> For these participants, accuracy was determined based on the
> production of the entire item (see Wambaugh et al., 2016 for details
> on these four participants). All items were probed in a
> multiple-baseline crossover design with at least 5 baseline
> observations. For participants with 20-item treatment lists, accuracy
> was determined based on the production of the target phoneme within
> the item. For participants with 10-item treatment lists, the entire
> item was scored (see Wambaugh et al., 2016 for details on these four
> participants).

| Variable    | Description                                                   |
|:-----------------|:-----------------------------------------------------|
| participant | de-identified participant ID                                  |
| condition   | probe schedule (blocked or random)                            |
| phoneme     | target_phoneme                                                |
| itemType    | item condition (treatment or generalization)                  |
| phase       | treatment phase                                               |
| session     | session number from Wambaugh 2017                             |
| item        | item identifier                                               |
| trials      | number of items in the list (per phoneme)                     |
| spt2017     | phase used to calcualte effect sizes in Wambaugh et al., 2017 |
| response    | accuracy of participant response                              |
| n_baselines | Number of baseline sessions                                   |

: Table 1. Variables and descriptions for study data from Wambaugh et
al., (2017)

## Setup

Before you get started, we're going to do this exercise as if you were
starting from scratch.

1.  In Rstudio, go to file, then select "new project". Choose "New
    Project", then "New Directory", then save the new project in a place
    that makes sense for you on your computer. Give the project a useful
    name.
2.  Create a folder for data and a folder for scripts
3.  In your finder (mac) or file explorer (windows), copy the
    "sound-production-treatment.csv" file from the data folder in the
    2022 workshop folder to the data folder in your new project.
4.  Create a new Rmarkdown script by going to file ---\> new file ---\>
    Rmarkdown.
5.  Change the title and author. Delete everything after the first code
    chunk.
6.  If you prefer, you can copy the remaining contents of this HTML
    Document to your new R markdown file by opening the Rmarkdown
    version of the file, and coping the content.

## Part 1.

For each question, save your code in the area provided below the
question.

**Q1.0 Load the tidyverse and here libraries and read in the
"sound-production-treatment.csv" dataframe**

```{r}
# Your code here
```

**Q1.1 Print a summary of each variable in the data. What do you notice
that might require further attention?**

```{r}
# Your code here

```

**Q1.2 Remove any NA values from the data, saving the data as a new
object**

```{r}
# Your code here

```

**Q1.3 Use `group_by` and `summarize` to create a new data frame that
reports on the mean accuracy of each participant for each session in the
baseline, treatment, and follow-up phases for treated items in the
blocked condition (use `filter` to include only these phases and items,
and condition).**

```{r}
# Your code here

```

**Q1.3 bonus: How can you keep the `phase` variable in this summary
data?**

```{r}
# Your code here

```

**Q1.4. The following code prints out a value from the dataset. Below
the code chunk, write a sentence that uses the value using inline R
code.**

```{r, eval = FALSE}
P10_bl_all <- df %>%
  # filter for participant 10, baseline phase, random condition, treated items
  filter(participant == "P10",
         phase == "baseline",
         condition == "random",
         itemType == "tx") %>%
  # get the mean accuracy across all 
  summarize(accuracy = mean(response)*100) %>%
  pull(accuracy)

# Write the following sentence below this chunk (in normal markdown text), but use inline R
# code to insert the P10_bl_all object's value. 

# Participant 10's baseline accuracy was ____%.
  
```

**Q1.4 Pick one participant and plot their performance across all
treatment phases.**

-   use the shape aesthetic for treatment phase
-   use the color aesthetic for item type
-   Add facet_wrap(\~condition, ncol = 1) to the final line of your plot
    to add a "facet" to plot the conditions separately.
-   label the y axis "Percent Correct"
-   Give the plot a title
-   Change the plot theme
-   Bonus #1: Move the legend below the plot
-   Bonus #2: Change the colors used in the plot

```{r}
# Your code here

```

## Part 2.

**Hopefully by now you've gotten a sense of what kind of variables are
in the data. In Part 2, we'd like you to explore some aspect of the data
that you find interesting. Can you wrangle the data and create a plot to
show the group that gets at a question you have about the data.**

-   Comment your code clearly so that you can explain what your code
    does

-   If your code requires multiple steps, use multiple code chunks and
    write a brief description of the goal of each chunk above it.

```{r}
# Your code here

```
